Dissecting the relationship between protein structure and sequence variation

### 2015-06-03 manuscript/voronoi/mnv.tex
MN update + new figures

    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp} and the values of the free parameters of CN and WCN that yield the strongest correlations are tabulated in Table \ref{tab:best_params}. Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of the long-range amino acid interactions, beyond the immediate neighborhood of the site, to the strengths of the observed correlations. The significance and impact of these non-local interactions however, cannot be deciphered solely from the two common definitions of packing density: CN \& WCN. In the following section we present a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability measures.

@@

    There is already an extensive body of literature on applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}. The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and the development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}. For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell corresponding to each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points. \\

@@

    In order to assess the prediction power of the site-specific characteristics derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra. We show in Appendix \ref{app:best_crd} that the choice of the geometric centers of the side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.  \\

    Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and with other Voronoi cell properties. In general, it is observed that the cell volume and surface area have the best prediction power compared to other cell characteristics, followed by the cell eccentricity, the total edge length, and the cell sphericitiy. \\

    The Voronoi cell characteristics are also strongly associated with each other. Although the cell volume and area are almost identically the best correlating variables with ER, the cell volume does not exhibit any significant independent correlation with ER once the cell area is controlled for. The median strength of the partial correlation of volume with ER, while controlling for area is centered at $\sim0.0$ (c.f., Figure \ref{fig:voronoi_ER_cond}). Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to represent almost the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. However, these cell characteristics are discrete (integer) quantities and in general have a limited range. \\

    Not shown here for brevity, almost identical results to the above are obtained were sequence entropy used in place of evolutionary rates, as defined by Eqn. \ref{eqn:shannon}. The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy. One potential caveat with the Voronoi tessellation of finite structures such as proteins is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure. We show in appendix \ref{app:edge_effects} that such effects are generally negligible in our results presented in this section. \\

    \subsection*{Effect of Long-Range Amino Acid Interactions on Sequence Variability}
    \label{sec:long_range}
    Figure \ref{fig:best_predictorER} compares the prediction power of each of the site-specific structural quantities about sequence evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes. Not shown here for brevity, similar results are also obtained for sequence entropy as the measure of sequence variability. For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the coordinates of the geometric centers of side chains. Notably, the quantity WCN outperforms all other structural quantities in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}. The better performance of WCN compared to local packing density as measured by the inverse of Voronoi cell volume may not be surprising, knowing that WCN by its definition in Eqn. \ref{eqn:wcn_pwrl} also takes into account the potential long-range interactions among amino acids in different regions of protein. \\

    In order to segregate the combined effects of long-range interactions from local packing density, the inverse of the Voronoi cell volume can be used as a maximally local measure of packing density. By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the effects long-range amino acid interactions beyond immediate neighbors -- on sequence variability. Figure \ref{fig:best_predictorER_given_volume} illustrates the partial correlation strengths of the same structural quantities as in Figure \ref{fig:best_predictorER}, while controlling for the Voronoi cell volume. It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman correlation strength $\rho$ is a rank correlation coefficient.

@@

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{best_structural_predictors_of_ER.pdf}
        \end{center}
        \caption{A comparison of the prediction power of five structural variables about site-specific evolutionary rates (ER). All structural quantities correlate positively with ER, with the exception of Weighted Contact Number (WCN) which correlates negatively. For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure. Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$. The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
        \label{fig:best_predictorER}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{best_structural_predictors_of_ER_given_VSCvolume.pdf}
        \end{center}
        \caption{A comparison of the prediction power of four structural variables (as in Figure \ref{fig:best_predictorER}) about site-specific evolutionary rates (ER), while controlling for the voronoi cell volume. All structural quantities correlate positively with ER on average, with the exception of Weighted Contact Number (WCN) which correlates negatively. For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure. Note that the Spearman's $\rho$ is insensitive to the use of inverse WCN in place of WCN. The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
        \label{fig:best_predictorER_given_volume}
    \end{figure}

>>>>

    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp} and the values of the free parameters of CN and WCN that yield the strongest correlations are tabulated in Table \ref{tab:best_params}. (XX a comparison of CN cutoff distance with average amino acid neighbor distance is needed here to quantify evidence for the following sentence. An additional figure may also be necessary plotting cutoff distance vs. average NN distance for each protein XX). Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of the long-range amino acid interactions, beyond the immediate neighborhood of the site, to the strengths of the observed correlations. The significance and impact of these non-local interactions on sequence evolution cannot be deciphered solely from the two common definitions of packing density: CN \& WCN. Therefore, we present, in the following section, a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability measures.

@@

    There is already an extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its relation to sequence \cite{richards_interpretation_1974, gerstein_volume_1994}. The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and the development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}. For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the Euclidean space into regions, called {\it cells}, such that the cell corresponding to each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points. \\

@@

    In order to assess the prediction power of the site-specific characteristics derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra. We show in Appendix \ref{app:best_crd} that the choice of the geometric centers of the side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in strongest correlations of Voronoi cell properties with site-specific sequence variability of proteins.  \\

    Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and with other Voronoi cell properties. In general, it is observed that the cell volume and surface area have the best predictive power compared to other cell characteristics, followed by the cell eccentricity, total edge length, and the cell's sphericitiy. \\

    The Voronoi cell characteristics are also strongly associated with each other. Although the cell volume and area are almost identically the best correlating variables with ER, the cell volume does not exhibit any significant independent correlation with ER once the cell area is controlled for. The median strength of the partial correlation of volume with ER, while controlling for area is centered at $\sim0.0$ (c.f., Figure \ref{fig:voronoi_ER_cond}). Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to represent almost the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. However, these cell characteristics are discrete (integer) quantities and in general have limited ranges. \\

    Not shown here for brevity, almost identical results to the above are obtained were sequence entropy used in place of evolutionary rates, as defined by Eqn. \ref{eqn:shannon}. The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range that is inherent in the definition of sequence entropy. One potential caveat with the Voronoi tessellation of finite structures, such as proteins, is the presence of {\it edge effects} in the properties of the cells that remain open, typically on the surface of the structure. We show in appendix \ref{app:edge_effects} that such effects are generally negligible in our results presented in this section. \\

    \subsection*{Effect of Long-Range Amino Acid Interactions on Sequence Evolution}
    \label{sec:long_range}
    Figure \ref{fig:best_predictorER} compares the prediction power of each of the site-specific structural quantities about sequence evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes. Not shown here for brevity, similar results are also obtained for sequence entropy as the measure of sequence variability. For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the coordinates of the geometric centers of side chains. Notably, the quantity WCN outperforms all other structural quantities in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}. The better performance of WCN compared to local packing density as measured by the inverse of Voronoi cell volume may not be surprising, knowing that WCN by its definition in Eqn. \ref{eqn:wcn_pwrl} also takes into account the potential long-range interactions among amino acids in different regions of protein. \\

    In order to segregate the combined effects of long-range interactions from local packing density, the inverse of the Voronoi cell volume can be used as a maximally local measure of packing density. By controlling for the Voronoi cell volume, one can then quantify the residual influence of WCN -- that is, the effects long-range amino acid interactions beyond immediate neighbors -- on sequence variability. Figure \ref{fig:best_predictorER_given_volume} illustrates the partial correlation strengths of the same structural quantities as in Figure \ref{fig:best_predictorER}, while controlling for the Voronoi cell volume. It should be noted that the strengths of the partial correlations are insensitive to whether the cell volume or its inverse is controlled for, since the Spearman correlation strength $\rho$ is a non-parametric rank correlation coefficient.

@@

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{best_structural_predictors_of_ER_limited.pdf}
        \end{center}
        \caption{A comparison of the prediction power of five structural variables about site-specific evolutionary rates (ER). All structural quantities correlate positively with ER, with the exception of Weighted Contact Number (WCN) which correlates negatively. For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure. Note that the Spearman's $\rho$ is a rank correlation coefficient, meaning that the use of inverse WCN only changes the sign and not the magnitude of $\rho$. The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
        \label{fig:best_predictorER}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{best_structural_predictors_of_ER_given_vol.pdf}
        \end{center}
        \caption{A comparison of the prediction power of four structural variables (as in Figure \ref{fig:best_predictorER}) about site-specific evolutionary rates (ER), while controlling for the voronoi cell volume. All structural quantities correlate positively with ER on average, with the exception of Weighted Contact Number (WCN) which correlates negatively. For better illustration however, the Spearman's correlation coefficient ($\rho$) of the inverse of WCN with ER are shown in the Figure. Note that the Spearman's $\rho$ is insensitive to the use of inverse WCN in place of WCN. The abbreviation {\it SC} refers to the use of average Side-Chain coordinates or average Side-Chain B factor wherever used, and {\it CA} refers to the use of backbone C$_\alpha$ atomic coordinates for representation of individual sites in proteins.}
        \label{fig:best_predictorER_given_volume}
    \end{figure}
